---
title:  "Reproducible Research - Project 1"
author: "analytics-moses"
date:   26-May-19
output: 
        md_document:
                variant: markdown_github
---
# Reproducible Research - Project 1

## Introduction
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a [Fitbit](), [Nike Fuelband](http://www.nike.com/us/en_us/c/nikeplus-fuelband), or [Jawbone Up](https://jawbone.com/up). These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behaviour, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

* Activity Monitoring Data Set: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

##Preliminary steps: Loading Prerequisite Packages and Downloading the Data Set
```{r}
library(ggplot2)
library(data.table)

if(!file.exists('activity.csv') ){
        if (!file.exists('repdata_data_activity.zip')){
              url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
              download.file(url, destfile = paste0(getwd(), '/repdata_data_activity.zip'), method = "curl") 
        }
    unzip('repdata_data_activity.zip')
}
```

1. Code for reading in the dataset and processing the data
```{r}
DT <- data.table::fread(input = "activity.csv")

```

## What is the mean total number of steps taken per day?
```{r}
#constructing a sub-set of DT data table with missing values (NA) present
columns <- "steps"

totalSteps <- DT[, c(lapply(.SD, sum, na.rm = FALSE)), .SDcols = columns, by = .(date)]

```

### Verifying that the Data Set which was input is correct:
```{r}
head(totalSteps,10)
```

2. Histogram of the total number of steps taken each day
```{r}
hist_plot <- ggplot(totalSteps, aes(steps)) +
        geom_histogram(fill = "darkgreen", binwidth = 500) +
                labs(title = "Total Daily Steps", x = "Steps", y = "Frequency (w/Binwidth 500)") 

plot(hist_plot)
```

3. Reporting Mean and Median number of steps taken each day:
```{r}
totalSteps[, .(`Mean # of Steps` = mean(steps, na.rm = TRUE), `Median # of Steps` = median(steps, na.rm = TRUE))]

```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

** Time series plot of the average number of steps taken:
```{r}
interval <- DT[, c(lapply(.SD, mean, na.rm = TRUE)), .SDcols = columns, by = list(interval)]

head(interval,10)

ggplot(interval, aes(x = interval , y = steps)) + geom_line(color="darkgreen", size=0.5) + labs(title = "Average Daily Steps", x = "5-min Interval", y = "Average Steps per Day")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
interval[steps == max(steps), list(`Maximum Intreval` = interval)]
```

## Imputing missing (NA) values
1.	Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
message("Total number of missing values (NAs) in the orig. Data Set: ", nrow(DT[is.na(steps),]))

```


2. Devise a strategy for filling in all of the missing values in the dataset. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r}
# Imputing the missing values (NA) in Data Set (DT) with Median values 
DT[is.na(steps), "steps"] <- DT[, c(lapply(.SD, median, na.rm = TRUE)), .SDcols = columns]

# verifying Data Set (DT) after imputing missing values
head(DT, 10)
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}

# saving the newly updated DT
data.table::fwrite(x = DT, file = "clean_set.csv", quote = FALSE)
```


4. a) Plotting a histogram for the total number of steps taken each day.  What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}

totalSteps_imp <- DT[, c(lapply(.SD, sum)), .SDcols = columns, by = list(date)] 

head(totalSteps_imp,20)
```

4.b) What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
ggplot(totalSteps_imp, aes(x = steps)) + geom_histogram(fill = "darkgreen", binwidth = 500) + labs(title = "Average Daily Steps", x = "Steps", y = "Frequency (w/Binwidth 500)")

```

4.c) Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment?

* Mean and Median # of Steps based on the imputed Data Set:
```{r}

# mean and median total number of steps taken per day
totalSteps_imp[, .(`Mean # of Steps` = mean(steps), `Median # of Steps` = median(steps))]

```

### Vis-a-vis Mean and Median # of steps WITHOUT missing values (NA) getting subsituted with Medians
```{r}
totalSteps[, .(`Mean # of Steps` = mean(steps, na.rm = TRUE), `Median # of Steps` = median(steps, na.rm = TRUE))]
```

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
# re-read DT from the imputed Data Set
DT <- read.csv('clean_set.csv')

# if 'date' is Sunday or Saturday, then it is a Weekned, else: Weekday
DT$DayType <-  ifelse(as.POSIXlt(DT$date)$wday %in% c(0,6), 'Weekend', 'Weekday')
```

2. Make a panel plot containing a time series plot (i.e.  type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
# create an aggregated table: group by interval, then DayType and output a 'mean' # of steps for e/grouping
aggrDT <- aggregate(steps ~ interval + DayType, data=DT, mean)

ggplot(aggrDT, aes(interval, steps)) + 
    geom_line(color="darkblue", size=0.5) + 
    facet_grid(DayType ~ .) +
    xlab("5-min Interval") + 
    ylab("Average Number of Steps")

```
